Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2018-11-01 anzjs2018/sections/04_implementation.tex
Clean up dir

The application consists of two components:
the first handles importing static GTFS data and constructing the network,
while the second implements the \rt models and prediction.
We chose \verb+Rcpp+ to develop the program,
which provides access to other R packages for data manipulation 
(notably \verb+RSQLite+ and \verb+dplyr+),
as well as the speed and memory management capabilities of \verb|C++|. 
The program is implemented in the R package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component
and assess its performance in \rt,
both with respect to timing and parameter estimation.

The general structure of the \rt component is shown below,
with the bold steps being those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Each time new data are received \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from the new data
    \item \textbf{Run particle filter on each vehicle to update or initialise state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles using combination of particle filter and network state
    \item Write ETAs to extended Google Protobuf binary file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarily concerned with ensuring each component of the program
is as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provides the memory management control necessary to make the application
viable in real-time.
We also make use of OpenMP for parallelisation,
since each vehicle is modelled independently so it is trivial to scale up without
worrying about thread safety.

The number of particles needed per vehicle 
depends on many factors,
so it is necessary to explore the performance of the application
with varying number of particles.
Application performance is also assessed
for a range of fixed model parameter values,
such as system noise and GPS error.
To enable comparisons, we implemented a simulated \rt environment
in which the same subset of real vehicle data from 8~October, 2018
could be analysed using a range of settings.
These simulations were carried out of a Virtual Machine 
with 8~cores and 32~Gb of memory, 
running Ubuntu~16.04 and using R~3.4.1.

@@

In each iteration, 
the timing of the various program components is recorded.
Since the number of vehicles travelling at any given time changes throughout the day,
we used the average timings over an off-peak 15~minute window.
Figure~\ref{fig:timings} shows the average timings for 
varying numbers of particles, $N$.

The most time consuming component is ETA writing,
which involves summarising the individual ETAs estimated for each particle.
While not discussed here this involves quantiles which use a sorting algorithm,
which has complexity $O(N \log N)$ 
which explains the slightly quadratic trend in the timings.

The next most intensive step is vehicle updating,
and takes about 5~seconds for 8000 particles.
The ETA prediction step only takes a few seconds,
although this may increase once a more comprehensive model is developed.
On our 8-core virtual machine the entire process is completed within 20~seconds
when 8000~particles are used.
In the next section we explore the effect of changing $N$
to allow finding the optimal number.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_timing.pdf}
    \caption{
        Application timings were recorded for each component of each
        iteration, and averages over a 15~minute window are shown here.
        The timings shown are \emph{wall time}, the actual time passed,
        not CPU time which would be approximately 8~times larger 
        (for using 8~cores).
        Steps that only involve particle calculations are approximately
        of complexity $O(N)$,
        however some steps are a little more complex, 
        so there is a slightly exponential trend.
    }
        % The timings for various parts of the application, and overall. %
        % The trend is slightly non-linear due to the complexity of the ETA generation step, %
        % which requires sorting of the particle ETAs to obtain quantile estimates.}
    \label{fig:timings}
\end{figure}

@@

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\epsilon$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
so we chose to use GPS error values $\epsilon \in \{1,2,3,5\}$. 
For the system noise we use values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to the average vehicle speed having a variance between 0.1 and 45~meters per second
over a period of 30~seconds.
For each of the combination of these parameter values, along with varying $N$,
several values were computed.
These were effective sample size, $N_\text{eff}$ calculated using (\ref{eq:neff});
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the variance of segment travel time estimates.

@@

From Figure~\ref{fig:degen_rate}, we see that $N_\text{eff}$ increases
with GPS error and decreases with system noise.
The degeneration rate decreases with GPS error and increases with $N$,
but is relatively unaffected by system noise.
Large GPS error affects the likelihood,
and gives more weight to particles farther from the observation
than does a smaller GPS error.
Conversely, increasing system noise spreads out the particle cloud,
so fewer particles will be near the vehicle.
Thus, these results are not surprising, 
but show that the model is working as expected.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
    \caption{
        The degeneration rate is the percentage of iterations where there were no
        plausible particles, so the vehicle was ``lost'' by the model.
        This decreases as GPS error and number of particles increase,
        but there is no visible effect of system noise.
        The error bars show $\pm 2$SE.
    }
    \label{fig:degen_rate}
\end{figure}

The last measure of performance is the relative variance of segment travel times.
This is calculated by computing the average variance of travel times over all simulations,
and then for each simulation, 
taking the average (over road segments) of the ratio between variance for that simulations
and variance for all simulations.
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
result in a higher relative variance,
while simulations with more particles have lower relative variance.
There is no visible effect of system noise on travel time variation.

The results of the simulations
demonstrate a tradeoff between particle filter performance
($N_\text{eff}$ and degeneration rate) and parameter estimation.
However, until the arrival time prediction model has been completed,
it is difficult to make decisions about the optimal values to use:
we are unable to say whether a higher variation of travel time estimates
is going to have a significant effect on the final ETAs,
particularly when compared to the uncertainty of forecasting road state.

>>>>

The application consists of two components:
the first handles importing static GTFS data and constructing the network,
while the second implements the \rt models and prediction.
We used Rcpp \citep{Rcpp}
to develop the program,
which provides access to other R packages for data manipulation 
as well as the speed and memory management capabilities of C++.
The program is implemented in the R package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component
and assess its performance 
both with respect to timing and parameter estimation.

The general structure of the \rt component is shown below,
with the bold steps being those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Each time new data are received \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from the new data
    \item \textbf{Run particle filter on each vehicle to update or initialise its state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles using combination of vehicle and network state
    \item Write ETAs to a file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarily concerned with ensuring each component of the program
was as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provided the memory management control necessary to make this possible.
We also made use of OpenMP for parallelisation,
since each vehicle is modelled independently so it is trivial to scale up without
issues with thread safety.

The number of particles needed per vehicle 
depends on many factors,
so it was necessary to explore the performance of the application
with varying number of particles.
Application performance is also assessed
for a range of fixed model parameter values,
such as system noise and GPS error.
To enable comparisons, we implemented a simulated \rt environment
in which the same subset of real vehicle data from 8~October, 2018
could be analysed using a range of settings.
These simulations were carried out of a Virtual Machine 
with 8~cores and 32~Gb of memory, 
running Ubuntu~16.04 and using R~3.4.1.

@@

In each iteration, 
the timing of the various program components was recorded.
Since the number of vehicles travelling at any given time changes throughout the day,
we used the average timings over an off-peak 15~minute window.
Figure~\ref{fig:timings} shows the average timings for 
varying numbers of particles, $N$.

The most time consuming component was ETA writing,
which involved summarising the individual ETAs estimated for each particle.
While not discussed here this involved computing quantiles which uses a sorting algorithm
with complexity $O(N \log N)$,
which explains the slightly quadratic trend in the timings.

The next most intensive step was vehicle updating,
which took about 5~seconds for 8000 particles.
The ETA prediction step only took a few seconds,
although this may increase once a more comprehensive model is developed.
On our 8-core virtual machine the entire process is completed within 20~seconds
when 8000~particles are used.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_timing.pdf}
    \caption{
        Application timings were recorded for each component of each
        iteration, and averages over a 15~minute window are shown here.
        The timings shown are \emph{wall time}, the actual time passed,
        not CPU time.
        Steps that only involve particle calculations are approximately
        of complexity $O(N)$,
        however some steps are a little more complex (writing ETAs involves computing quantiles), 
        so there is a slightly exponential trend.
    }
        % The timings for various parts of the application, and overall. %
        % The trend is slightly non-linear due to the complexity of the ETA generation step, %
        % which requires sorting of the particle ETAs to obtain quantile estimates.}
    \label{fig:timings}
\end{figure}

@@

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\epsilon$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error. 
For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to an average vehicle speed variance of between 0.1 and 45~meters per second
over a period of 30~seconds.
For each of combination of these parameter values, including varying $N$,
the following values were calculated:
effective sample size, $N_\text{eff}$;
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the (relative) variance of segment travel time estimates.

@@

Figure~\ref{fig:neff} shows that $N_\text{eff}$ increased
with GPS error and decreased with system noise,
while remaining unaffected by changing $N$.
The degeneration rate decreased with GPS error and increased with $N$,
and was unaffected by system noise,
as seen in Figure~\ref{fig:degen_rate}.
Large GPS error affects the likelihood,
giving more weight to particles farther from the observation
than does a smaller GPS error.
Conversely, increasing system noise spreads out the particle cloud,
so fewer particles will be near the vehicle.
Thus, these results are not surprising, 
but show that the model is working as expected.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
    \caption{
        The degeneration rate is the percentage of iterations where there were no
        plausible particles, so the vehicle was ``lost'' by the model.
        This decreases as GPS error and number of particles increase,
        but there is no visible effect of system noise.
        Note, however, that some of these can be attributed to invalid data,
        for example if the driver has entered the wrong trip direction.
        The error bars show $\pm 2$SE.
    }
    \label{fig:degen_rate}
\end{figure}

The last measure of performance is the relative variance of segment travel times.
We used relative variance as each segment has a very different travel time distribution,
so to compare the variability between simulations we needed to 
compute the average variance of travel times over all simulations,
taking the average (over road segments) of the ratio between variance for
each individual simulation and variance over all simulations.
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
resulted in a higher relative variance,
while simulations with more particles had lower relative variance.
There was no visible effect of system noise on the variability of travel times.

The results of the simulations
demonstrate a tradeoff between particle filter performance
($N_\text{eff}$ and degeneration rate) and parameter estimation.
However, until the arrival time prediction model has been completed,
it is difficult to make decisions about the optimal values to use:
we are unable to say whether a higher variation of travel time estimates
is going to have a significant effect on the final ETAs,
particularly when compared to the uncertainty of forecasting road state,
or whether reducing the degeneration rate, 
or increasing $N_\text{eff}$, would be more advantageous.

### 2018-11-01 anzjs2018/sections/06_discuss.tex
Clean up dir

In this paper we have described an approach to transit vehicle modelling
that enables the \rt estimation of road state from vehicle position data,
which can later be used to make arrival time predictions
that account for \rt traffic conditions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modelled,
with future work planned to extend the model to handle physical intersections,
such as traffic lights.
From the particle filter,
vehicle travel times along roads are estimated which are in turn used 
in the modelling of the road network state.

The network model presented here is simple,
but provides the foundation for future work to
develop a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
This in turn will provide the basis for an arrival time prediction model
that combines \rt vehicle state with travel time forecasts
and provide commuters with reliable ETAs.
These will likely consist of a point estimate along with a prediction interval,
as it is impossible given the available data to accurately predict arrival time
due to traffic lights, unknown dwell times at intermediate bus stops,
and many other factors.

>>>>

In this paper we described an approach to transit vehicle modelling
that enables the \rt estimation of road state from vehicle position data,
which in the future will be used to make arrival time predictions
that account for \rt traffic conditions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modelled,
with future work planned to extend the model to handle physical intersections,
such as traffic lights.
From the particle filter,
vehicle travel times along roads are estimated which are in turn used 
in the modelling of the road network state.

The network model presented here is simple,
but provides the foundation for future work to
develop a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
This in turn will provide the basis for an arrival time prediction model
that combines \rt vehicle state with travel time forecasts
and provide commuters with what will hopefully be reliable ETAs.
These will likely consist of a point estimate along with a prediction interval,
as it is impossible given the available data to accurately predict arrival time
due to traffic lights, unknown dwell times at intermediate bus stops,
and many other factors.
